Executive Constraints and Civil Conflict Onset

### 2020-08-30 paper.Rmd
Manuscript edits

---
title: Executive Constraints and Civil Conflict Onset
author:
    name: J.L.A. Krusell
    affiliation: Department of Peace and Conflict Research
    supervisor: Kristine Eck
    degree: "Master's Thesis"
date: "Spring 2020"
titlelogo: |
    ```{=latex}
    assets/UU_logo_CMYK-eps-converted-to.pdf
    ```
wordcount: 14300
suppress-bibliography: false
bibliography: library.bib
csl: assets/chicago-author-date.csl
link-citations: true
documentclass: scrreprt
classoption: paper=letter
geometry:
    - top=30mm
    - heightrounded
fontsize: 12pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
    - \usepackage[font={small}]{caption}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--template", "assets/thesis_template.latex",
                      "--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: Lorem ipsum
---

@@

Since the end of World War II, civil conflict has been the most
prevalent form of state-based conflict. Of the
`r fmt(total_conflicts)` conflicts recorded by the UCDP/PRIO Armed
Conflict Database v19.1 [@Gleditsch2002; @Pettersson2019] for the
years covering 1946 until 2018, `r fmt(intrastate_conflicts)` of
those, or approximately `r fmt(intrastate_percent, 1)` percent, are
classified as intrastate. These civil conflicts are distributed across
`r fmt(n_states)` countries, more than half of which having
experienced at least one major civil war with 1,000 battle deaths at a
minimum.  Where data is available, civil conflicts are also
empirically among the deadliest. Out of `r fmt(total_deaths)`
cumulative battles deaths from organized violence recorded
since 1989 [@Pettersson2019; @Eck2007; @Sundberg2012],
`r fmt(civil_deaths)`, or `r fmt(civil_deaths_percent, 1)` percent, are
the result of civil conflict.

Given the impact of civil conflict and the increasing availability of
conflict related data, the last several decades has witnessed a
veritable explosion in the number of quantitative studies
investigating the causes of civil conflict [@Blattman2017]. Scholars
have proposed numerous theories ranging from motivation-based
arguments --- such as, for example, the role of horizontal
inequalities and ethnic identities [@Cederman2011] --- to
opportunity-based explanations centered on economic factors
[@Collier2004; @Collier2009] as well as the political and structural
conditions underlying state weakness [@Hegre2001; @Fearon2003]. This
latter strain of research in particular has spurred extensive work
into examining the distinguishing features of weak states and whether
political institutions play a significant role in the risk of civil
conflict. Several studies have argued in favor of institutions by
investigating, for example, aspects of governance, institutional
quality, and opportunities for political participation [@Fjelde2009;
@Taydas2010; @Hegre2015a; @Bartusevic2018].

This paper extends that work by taking a disaggregated approach to
studying political institutions by focusing specifically on the effect
of executive constraints on the risk of civil conflict onset. The
motivation to focus on executive constraints is two-fold: first, the
executive is often times one of the key decision-makers in a state's
regime. Institutions that provide a check against the types of actions
which may exacerbate the risk of rebellion should be relevant to the
study of civil conflict. Such an approach can further our
understanding of how specific political institutions may shape the
incentives for violence by affecting state behaviour. Second, the link
between executive constraints is relatively understudied. The previous
literature that exists is fairly limited and contradictory in terms of
results.

The principle finding of this paper is that there is tentative
evidence that executive constraints do matter and that there is a
negative relationship between the latent level of constraints on the
executive and the risk of civil conflict onset. However, this
relationship largely disappears for low-income country-years when an
interaction term with GDP per capita is included. If the primary
theoretical mechanism involves the impact of executive constraints on
rebel mobilization --- an unchecked executive may be more likely to
engage in indiscriminate violence which could lead to a loss of
legitimacy as well as the ability of rebel groups to offer local
security guarantees to attract recruits --- then the incentives for
violence in low-income country-years may be so great that it does not
matter how constrained the executive may or may not be.

@@

# Theoretical Overview {-}

## Institutions and Civil Conflict

This paper takes as a starting premise a definition of institutions as
being "systems of established and embedded social rules that structure
social interactions" [@Hodgson2006]. More informally, we can consider
them to be the "rules of the game", encompassing for example codified
laws, norms, and shared understandings, all of which can constrain or
enable behaviour by influencing the perceptions and preferences of
actors through incentives and deterrents.[^institutions] Furthermore,
over time, the habits formed from these systems create expectations of
stable patterns of behaviour thereby further shaping the interactions
between actors by reducing uncertainty [@North1990]. Although there is
a wide variety of institutions, the present focus will be on the type
of institutions which govern political behaviour.

@@

This exposition is relevant to the study of civil conflict for both
motivation-based and opportunity-based explanatory accounts. To the
extent that the onset of civil conflict is a longer process featuring
waves of escalation and de-escalation rather than a single one-off
event [@Florea2012; @Sambanis2004b], institutions are a key influence
on the constitutive interactions between state and non-state
actors. For example, they can shift the opportunity costs of violence
altering the mobilization capacity of would be rebel
groups. Institutions can also determine the extent to which the state
is seen as accountable to and representative of the larger population,
thereby influencing the legitimacy of anti-government violence.  And,
their quality can shape the capacity of the state to suppress or
co-opt potential opposition. Although the preceding examples are far
from forming an exhaustive list, they illustrate the number of
different ways in which institutions can enter into theoretical
discussions about civil conflict.

Unsurprisingly, insofar as the role of institutions on the likelihood
of civil conflict has been examined by previous literature the focus
has largely been on the broader notion of democracy, which at a
minimum is a system regulating access to political power through
popular representation determined by free and competitive elections
[@Schumpeter1962]. Democracy serves as a proxy for a number of
normatively desirable institutions and has been hypothesised to have a
pacifying domestic effect according to two predominant mechanisms. The
first is centered on the concept of relative deprivation, or
inequalities in value expectations and value capabilities, as a
primary source of discontent that can motivate collective violence
[@Gurr1970]. [^gurr] Democracy is thought to decrease the amount of
political discontent by providing avenues for free expression and
adjudicating group conflicts through non-violent institutional means
--- for example, through the regular occurrence of free and fair
elections [@Hegre2014]. The second mechanism assumes that violent
conflict is costly and that there exists a bargained solution
preferable to fighting, assuming that actors are rational and able to
internalize the cost of violence [@Fearon1995]. One of the reasons why
a bargained solution may be impossible is the risk of future shifts in
the distribution of power. More specifically, a future change in favor
of the relative strength of one of the parties can provide an
incentive to later renege. This influences the present degree to which
concessions can be offered and parties are willing to credibly commit
[@Powell2006]. Within the domestic sphere, democracy offers a possible
solution by formalising the regular contest of power with full
enfranchisement, thereby providing a *de jure* guarantee for the
balance of power [@Origins2006], and should, likewise, decrease the
likelihood of civil conflict.

[^gurr]: Gurr explains value expectations as being that which we think
    we are entitled to and value capabilities as that which we think
    we are capable of attaining. As Gurr then states, relative
    deprivation is the perceived discrepancy between the two. The link
    to collective violence, however, is conditional on the discontent
    being politicized, by which it is meant that the source of the
    discrepancy is identified and a normative and utilitarian
    justification is given for violent action, before finally being
    actualized.  Quantitative studies tend to focus on the first
    stage, the propensity for political regimes to generate
    discontent, given the difficultly of measuring the variables that
    may affect politicisation and actualisation.

Despite the theoretical basis linking democracy to reduced risks for
civil conflict, the empirical results tend to be mixed. @Hegre2001,
for example, find that democracies are at a lower risk for civil
conflict; however, no more so than fully authoritarian
states. Instead, anocratic regimes in between democracy and
authoritarianism are considered to be the most prone. The explanatory
logic is that both democratic and fully authoritarian states tend to
be consolidated regimes. In the case of the former, this means the
presence of institutions conducive to conflict resolution such that
the opportunity for rebellion may be present, given the availability
of civil liberties conducive to organizing, but there is a notable
lack in motivation for collective violence. For the latter,
authoritarian regimes may produce a significant amount of political
discontent; however, the opportunity is lacking due to the state's
capacity and willingness to violently repress dissent. Anocratic
regimes are by their very nature transitory, partly open and partly
repressive, encouraging political discontent against the state without
the capacity to either accommodate or suppress opposition.

@@

Taken together, this recent literature on civil conflict onset points
to the ongoing importance of political institutions. There remains,
however, significant work to be done identifying the causal mechanisms
and effects of specific institutions, especially in terms of further
disaggregating the concept of democracy beyond ever finer
classifications of regime type [@ari2016]. Furthermore, to the extent
that previous research has mostly focused on the effects of
institutions on would-be rebel motivation and/or political opportunity
structures, there is also a gap in terms of empirical analyses on how
institutions also shape the behaviour of the state in the lead up to a
civil conflict. Even in strictly economic explanations of civil
conflict where explanatory primacy is given to either greed-based
motivation or economic conditions conducive for rebellion, the state
is a key actor. How a state responds to dissent structures the
opportunity costs of violence as well as the salience of calls to
mobilization by rebel leaders. This latter point is especially
relevant in linking regime legitimacy to conflict since public
perception is shaped not simply by a regime's formal structure, but by
its conduct in repeated interactions with the public. To this end, it
is imperative to examine the types of institutions which exacerbate
the risk of conflict onset by enabling or constraining certain actions
by the state.

There are of course a large variety of political institutions whose
presence and quality can influence the behaviour of the state.  Given
the importance of the executive as a central authority, however,
especially in times of political crisis, institutions which affect the
executive's decision making should be pertinent. Chief among these is
the idea of executive constraints --- more formally, the institutional
constraints by other government bodies on the executive's ability to
abuse power and act unilaterally. Executive constraints can be
considered a subset of horizontal accountability and imply the
existence of state agents --- such as the legislature or the judiciary
--- with both the legal mandate and willingness to oversee or sanction
the executive thus serving as deterrents to the forms of
authoritarian excesses in executive power which may be linked to civil
conflict. [^ODonnell]

@@

The final motivation for examining executive constraints is that in
terms of previous research, the empirical evidence is
conflicted. @Getmansky2013, for example, offers a theory of executive
constraints playing a role in risk aversion in democratic polities;
however, they find their measure to be insignificant in predicting the
outbreak of insurgencies. On the other hand, @Fjelde2009 with a theory
tying executive constraints to public goods provisioning reports a
significant interaction term with public spending. Both cases use the
ordinal measure of executive constraints form Polity IV
[@Marshall2019]. This leaves open further work on additional
theoretical mechanisms and analyses using other datasets.

## Executive Constraints

@@

The final pathway involves the impact of executive constraints on
shaping the ability of rebel groups to recruit new members. In order
to mount an effective challenge to the state, rebel leaders need to be
able to mobilize fighters while avoiding defections. However,
participating in an armed rebellion is a costly endeavour. Non-elites
face both the immediate risk of fighting as well as the possibility of
future reprisals by the government. Furthermore, given that most civil
conflicts are protracted events against an often times much stronger
opponent the short-term utility from participation is relatively
low. Thus, rebel leaders must be able to offer proportionate selective
benefits to offset the associated costs of fighting
[@Mason1996]. These benefits can include not only economic payoffs,
but also protection from government violence. When the government
engages in violent repression, especially in the form of
indiscriminate violence, the cost of non-participation is
raised. Rebel groups can provide local security guarantees against
government reprisals such that the opportunity cost of fighting is
lowered, thereby incentivizing participation [@Regan2005]. However,
this is predicated on the government lacking the capacity to fully
eradicate any rebel group. Where that capacity exists the state is
free to engage in reprisals.

It is at this point where the concept of executive constraints then
enters into the calculus. Leaders in nondemocratic states are more
likely to engage in violent repression as a response to internal
dissent [@Davenport2007a]. Executive constraints provide an
accountability mechanism deterring state violence by the leader
against their own citizens. Thus, where there are strong and effective
constraints on the executive, the opportunity cost for joining the
rebel side should be higher. Rebel leaders will then have to rely more
heavily on economic selective benefits [@Regan2005], creating a higher
barrier for mobilization and decreasing the risk of civil conflict
onset. It should be noted though, that where this relationship may not
hold is in situations of low intensity insurgencies involving
predatory rebel groups solely motivated by rent-seeking. The threshold
for mobilization in such scenarios is significantly lower than in full
scale civil wars, meaning that rebel groups can initiate a conflict
regardless of the current level of state repression.

Taken together, none of the aforementioned mechanisms are mutually
exclusive, and it is beyond the scope of the present paper to properly
adjudicate between them; however, they do converge on a single
testable proposition. Namely that higher levels of executive
constraints should be associated with a lowered risk of civil conflict
onset. This proposition can be further extended: if state weakness is a
primary driver of civil conflict and the relative capacity of the
state is effected by the levels of executive constraints, as depicted
in the first theoretical pathway regarding executive looting, then
state capacity could have a mediating affect on the relationship
between executive constraints and conflict onset. Another extension
would be if executive constraints affect the mobilization potential
for rebel groups, then there could also be a confounding effect from
the relative threshold for mobilization. Finally, the relationship
could still be confounded by the overall opportunity for rebellion ---
the degree of executive constraints may not be particularly relevant
in relatively stable countries lacking the conditions conducive to
carrying out a rebellion. Regardless of conditionalities, contrary
results would indicate either that executive constraints are
indeterminate to the risk of civil conflict or possibly that there is
even a positive relationship since constrained executives may be less
able to effectively deal with violent challenges due to risk aversion
to violence and/or from the slowdown of having to seek approval from a
plurality of stakeholders before committing to decisive action.

One final caveat, hitherto the relationship of executive constraints
to the notion of vertical accountability --- constraints on leaders by
free and open multiparty elections as well as an independent and
critical media --- has been left unspecified. This is not to diminish
the role of elections or the media; indeed, vertical accountability
can plausibly play a role in all of the mentioned mechanisms. For
example, studies on state repression have emphasised the importance of
non-elites in being able to vote elected offices out of office as an
effective deterrent against negative sanctioning by leaders
[@BuenoDeMesquita2005; @Davenport2007]. But elections occur only
intermittently and it is not an automatic given that voters will
punish an incumbent, especially when that repression is against an
oppressed minority community. Furthermore, while the media can expose
wrongdoing by government officials, legal punishment can only flow
through government bodies [@ODonnell1998]. Thus, executive constraints
are interesting in their own right.

@@

I model civil conflict onset as a multilevel regression where onset is
a function of a set of predictors, including executive constraints and
an array of control variables, together with several random
effects. The units of analysis are all independent country-years from
1952 until 2013, as determined by data availability, with *I*
countries indexed by $i = 1 \dotsc I$ and *T* years indexed by $t = 1
\dotsc T$. The dependent variable, conflict onset, is a Bernoulli
distributed binary variable where 1 denotes the onset of at least one
new civil conflict in a given country-year. The chosen link function
transforming the linear regression output into the probability of
conflict onset, $\nu_{it}$, is the inverse logit such that $\nu_{it}
\in [0, 1]$. [^invlogit]

[^invlogit]: Technically, the inverse logit transforms values to
    the open interval, $(0, 1)$, but due to overflow/underflow in
    floating point calculations by computers the endpoints can be
    observed.

@@

where **X** is the predictors, $\beta$ the vector of regression
slopes, and $\Zeta$ the *K* sets of varying intercepts indexed by
$j_k$. All predictors are lagged by one year to control for time
endogeneity and standardized by subtracting the mean and dividing by
the standard deviation to aid in computation as well as comparison of
the posterior $\beta$ estimates.

For the present analysis there are three sets of varying intercepts,
or random effects, included in the model. This includes two partially
pooled grouping factors, year and country, to account for
cross-sectional unmodeled heterogeneity between countries and over
time [@Shor2007], and the number of peace years since the last
civil conflict incidence per country to control for the lingering
endogenous effects of violence on the likelihood that additional
conflict breaks
out.

@@

TODO: grab this programmatically

>>>>

---
title: Executive Constraints and Civil Conflict Onset
author:
    name: J.L.A. Krusell
    affiliation: Department of Peace and Conflict Research
    supervisor: Kristine Eck
    degree: "Master's Thesis"
date: "Spring 2020"
titlelogo: |
    ```{=latex}
    assets/UU_logo_CMYK-eps-converted-to.pdf
    ```
wordcount: 15000
suppress-bibliography: false
bibliography: library.bib
csl: assets/chicago-author-date.csl
link-citations: true
documentclass: scrreprt
classoption: paper=letter
geometry:
    - top=30mm
    - heightrounded
fontsize: 12pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
    - \usepackage[font={small}]{caption}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--template", "assets/thesis_template.latex",
                      "--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: |
    **Abstract**: Do institutional constraints on a regime's executive decrease the
    likelihood of civil conflict onset? An unconstrained executive is free
    to loot state resources, undermining the state's capacity to
    effectively deal with nascent rebellions, and may be more likely to
    engage in violent repression, especially in the form of indiscriminate
    violence. This can encourage political grievances, lead to a loss of
    legitimacy for the regime, and provide the opportunity for rebel
    groups to attract new members by offering protection against
    government violence. Furthermore, the lack of guaranteed checks to
    executive power may incentivise actors to avoid bargained solutions if
    they fear future extralegal retributions. Taken together, where there
    is a lack of effective constraints against the executive there may be
    an increased risk for the onset of civil conflict. To test this
    proposition I employ a bayesian latent variable model which estimates
    executive constraint as a latent factor derived from several manifest
    variables and relates it to a binary measure of civil conflict onset
    through a logistic regression.  The primary finding is that there is a
    negative relationship such that the predicted probability for civil
    conflict is lower where there exists higher levels of executive
    constraints. However, this is conditional on the level of GDP per
    capita; for low-income country-years the relationship between
    executive constraints and civil conflict onset is indeterminate
    possibly because it is easier to recruit and mobilize fighters
    regardless of the overall level of executive constraints. The issue of
    poor model fit means that the presented results should be considered
    tentative at best.
---

@@

Since the end of World War II, civil conflict has been the most
prevalent form of state-based conflict. Of the
`r fmt(total_conflicts)` conflicts recorded by the UCDP/PRIO Armed
Conflict Database v19.1 [@Gleditsch2002; @Pettersson2019] for the
years covering 1946 until 2018, `r fmt(intrastate_conflicts)` of
those, or approximately `r fmt(intrastate_percent, 1)` percent, are
classified as intrastate. These civil conflicts are distributed across
`r fmt(n_states)` countries, more than half of which having
experienced at least one major civil war with 1,000 battle deaths at a
minimum.  Where data is available, civil conflicts are also
empirically among the deadliest. Out of `r fmt(total_deaths)`
cumulative battles deaths from organized violence recorded since 1989
[@Pettersson2019; @Eck2007; @Sundberg2012], `r
fmt(civil_deaths_percent, 1)` percent are the result of civil
conflict.

Given the impact of civil conflict and the increasing availability of
conflict related data, the last several decades have witnessed a
veritable explosion in the number of quantitative studies
investigating the causes of civil conflict [@Blattman2017]. Scholars
have proposed numerous theories ranging from motivation-based
arguments --- such as, for example, the role of horizontal
inequalities and ethnic identities [@Cederman2011] --- to
opportunity-based explanations centered on economic factors
[@Collier2004; @Collier2009] as well as the political and structural
conditions underlying state weakness [@Hegre2001; @Fearon2003]. This
latter strain of research in particular has spurred extensive work
into examining the distinguishing features of weak states and whether
political institutions play a significant role in the risk of civil
conflict. Several studies have argued in favor of institutions and
have investigated, for example, aspects of governance, institutional
quality, and opportunities for political participation as factors
influencing the risk of civil conflict [@Fjelde2009; @Taydas2010;
@Hegre2015a; @Bartusevic2018].

This paper extends that work by taking a disaggregated approach to
studying political institutions by focusing specifically on the effect
of executive constraints on the risk of civil conflict onset. The
motivation to focus on executive constraints is two-fold: first, the
executive is often times one of the key decision-makers in a state's
regime. Institutions that provide a check against the types of actions
which may exacerbate the risk of rebellion should be relevant to the
study of civil conflict. Such an approach can further our
understanding of how specific political institutions may shape the
incentives for violence by affecting state behaviour. Second, the link
between executive constraints and civil conflict is relatively
understudied. The previous literature that exists is fairly limited
and contradictory in terms of results.

The principle finding of this paper is that there is tentative
evidence that executive constraints do matter and that there is a
negative relationship between the latent level of constraints on the
executive and the risk of civil conflict onset. However, this
relationship largely disappears for low-income country-years when an
interaction term between executive constraints and gross domestic
product (GDP) per capita is included. If the primary theoretical
mechanism involves the impact of executive constraints on rebel
mobilization --- an unchecked executive may be more likely to engage
in indiscriminate violence which could lead to a loss of legitimacy as
well as the ability of rebel groups to offer local security guarantees
to attract recruits --- then the incentives for violence in low-income
country-years due to say economic factors may be so great that it does
not matter how constrained the executive may or may not be.

@@

# Theoretical Background {-}

## Political Institutions and Civil Conflict

This paper takes as a starting premise a definition of institutions as
being "systems of established and embedded social rules that structure
social interactions" [@Hodgson2006]. More informally, we can consider
them to be the "rules of the game", encompassing, for example, codified
laws, norms, and shared understandings, all of which can constrain or
enable behaviour by influencing the perceptions and preferences of
actors through incentives and deterrents.[^institutions] Furthermore,
over time, the habits formed from these systems create expectations of
stable patterns of behaviour thereby further shaping the interactions
between actors by reducing uncertainty [@North1990]. Although there is
a wide variety of institutions, the present focus will be on the type
of institutions which govern political behaviour.

@@

This exposition is relevant to the study of civil conflict for both
motivation-based and opportunity-based explanatory accounts. To the
extent that the onset of civil conflict is a longer process featuring
waves of escalation and de-escalation rather than a single one-off
event [@Florea2012; @Sambanis2004b], institutions are a key influence
on the constitutive interactions between state and non-state
actors. For example, they can shift the opportunity costs of violence
altering the mobilization capacity of would be rebel
groups. Institutions can also determine the extent to which the state
is seen as accountable to and representative of the larger population,
thereby influencing the legitimacy of anti-government violence. In
addition, institutional quality can shape the capacity of the state to
suppress or co-opt potential opposition. Although the preceding
examples are far from forming an exhaustive list, they illustrate the
number of different ways in which institutions can enter into
theoretical discussions about civil conflict.

Unsurprisingly, insofar as the role of institutions in shaping the
likelihood of civil conflict has been examined by previous
literature, the focus has largely been on the broader notion of
democracy, which at a minimum is a system regulating access to
political power through popular representation determined by free and
competitive elections [@Schumpeter1962]. Democracy serves as a proxy
for a number of normatively desirable institutions and has been
hypothesised to have a pacifying domestic effect according to two
predominant mechanisms. The first is centered on the concept of
relative deprivation, or inequalities in value expectations and value
capabilities, as a primary source of discontent that can motivate
collective violence [@Gurr1970]. [^gurr] Democracy is thought to
decrease the amount of political discontent by providing avenues for
free expression and adjudicating group conflicts through non-violent
institutional means --- for example, through the regular occurrence of
free and fair elections [@Hegre2014]. The second mechanism assumes
that violent conflict is costly and that there exists a bargained
solution preferable to fighting, assuming that actors are rational and
able to internalize the cost of violence [@Fearon1995]. One of the
reasons why a bargained solution may be impossible is the risk of
future shifts in the distribution of power. More specifically, a
future change in favor of the relative strength of one of the parties
can provide an incentive to later renege. This influences the present
degree to which concessions can be offered and parties are willing to
credibly commit [@Powell2006]. Within the domestic sphere, democracy
offers a possible solution by formalising the regular contest of power
with full enfranchisement, thereby providing a *de jure* guarantee for
the balance of power [@Origins2006], and should, likewise, decrease
the likelihood of civil conflict.

[^gurr]: Gurr explains value expectations as being that which we think
    we are entitled to and value capabilities as that which we think
    we are capable of attaining. As Gurr then states, relative
    deprivation is the perceived discrepancy between the two. The link
    to collective violence, however, is conditional on the discontent
    being politicized, by which it is meant that the source of the
    discrepancy is identified and a normative and utilitarian
    justification is given for violent action, before finally being
    actualized.  Quantitative studies tend to focus on the first
    stage, the propensity for political regimes to generate
    discontent, given the difficultly of measuring the factors that
    may affect politicisation and actualisation.

Despite the theoretical basis linking democracy to reduced risks of
civil conflict, the empirical results tend to be mixed. @Hegre2001,
for example, find that democracies are at a lower risk for civil
conflict; however, no more so than fully authoritarian
states. Instead, anocratic regimes in between democracy and
authoritarianism are considered to be the most prone to civil conflict
onset. The explanatory logic is that both democratic and fully
authoritarian states tend to be consolidated regimes. In the case of
the former, this means the presence of institutions conducive to
conflict resolution such that the opportunity for rebellion may be
present, given the availability of civil liberties conducive to
organizing, but there is a notable lack in motivation for collective
violence. For the latter, authoritarian regimes may produce a
significant amount of political discontent; however, the opportunity
is lacking due to the state's capacity and willingness to violently
repress dissent. Anocratic regimes are by their very nature
transitory, partly open and partly repressive, encouraging political
discontent against the state without the capacity to either
accommodate or suppress opposition.

@@

Taken together, this recent literature on civil conflict onset points
to the ongoing importance of political institutions. There remains,
however, significant work to be done identifying the causal mechanisms
and effects of specific institutions, especially in terms of further
disaggregating the concept of democracy beyond ever finer
classifications of regime type [@ari2016]. Furthermore, to the extent
that previous research has mostly focused on the effects of
institutions on would-be rebel motivation and/or political opportunity
structures, there is also a gap in terms of empirical analyses on how
institutions also shape the behaviour of the state in the lead up to a
civil conflict. Even in strictly economic explanations of civil
conflict where explanatory primacy is given to either greed-based
motivation or economic conditions conducive to rebellion, the state
is a key actor. How a state responds to dissent structures the
opportunity costs of violence as well as the salience of calls to
mobilization by rebel leaders. This latter point is especially
relevant in linking regime legitimacy to conflict since public
perception is shaped not simply by a regime's formal structure, but by
its conduct in repeated interactions with the public. To this end, it
is imperative to examine the types of institutions which exacerbate
the risk of conflict onset by enabling or constraining certain actions
by the state.

There are of course a large variety of political institutions for
which the presence and quality can influence the behaviour of the
state.  Given the importance of the executive as a central authority,
however, especially in times of political crisis, institutions which
affect the executive's decision making should be pertinent. Chief
among these is the idea of executive constraints --- more formally,
the institutional constraints by other government bodies on the
executive's ability to abuse power and act unilaterally. Executive
constraints can be considered a subset of horizontal accountability
and imply the existence of state agents --- such as the legislature or
the judiciary --- with both the legal mandate and willingness to
oversee or sanction the executive thus serving as deterrents to the
forms of authoritarian excesses in executive power which may be linked
to civil conflict. [^ODonnell]

@@

The final motivation for examining executive constraints is that in
terms of previous research, the empirical evidence is
conflicted. @Getmansky2013, for example, offers a theory of executive
constraints playing a role in risk aversion in democratic polities.
When the executive is constrained by other government institutions a
regime may be unwilling to resort to force against a growing
insurgency and may instead seek a peaceful solution to avoid
fighting.[^weakness] They use a dichotomized measure of XCONST,
"Constraint on Chief Executive", from Polity IV [@Marshall2019] in
order to test this theory, finding their measure to be insignificant
in predicting the outbreak of insurgencies.

[^weakness]: If the theory were to hold, it is unclear if risk aversion
    really translates to a lower incidence rate of civil conflict
    since such regimes may be subject to opportunistic rebel groups
    ready to exploit precisely those weaknesses. In other words,
    democratic regimes may actively attempt to avoid armed conflict;
    however, they can still be forced into such situations by rebel
    groups.

In another study, @Fjelde2009 put forth a theory tying executive
constraints to public provisioning. Increased government expenditures
in the presence of strong institutional checks against executive
corruption is thought to decrease the risk of civil conflict by
reducing potential grievances against the state. In contrast to
Getmansky, the authors find a significant negative effect for an
interaction term between executive constraints and government
expenditures when also using XCONST from Polity IV, although the
regression slope for the executive constraints term is insignificant.

In both studies, using an ordinal measure, and in the case of
@Getmansky2013 a binary measure, for executive constraints risks
losing a significant amount of information concerning the varying
degrees of effectiveness in the constraints against executives among
different countries. A possible improvement would be the use of a
continuous measure to be more able to capture the quality of
constraints and better adjudicate whether there exists a measurable
effect on civil conflict onset.

## Executive Constraints and Civil Conflict

@@

The final pathway involves the impact of executive constraints on
shaping the ability of rebel groups to recruit new members. In order
to mount an effective challenge to the state, rebel leaders need to be
able to mobilize fighters while avoiding defections. However,
participating in an armed rebellion is a costly endeavour. Non-elites
face both the immediate risk of fighting as well as the possibility of
future reprisals by the government. Furthermore, given that most civil
conflicts are protracted events against an often times much stronger
opponent, the short-term utility from participation is relatively
low. Thus, rebel leaders must be able to offer proportionate selective
benefits to offset the associated costs of fighting
[@Mason1996]. These benefits can include not only economic payoffs,
but also protection from government violence. When the government
engages in violent repression, especially in the form of
indiscriminate violence, the cost of non-participation is
raised. Rebel groups can provide local security guarantees against
government reprisals such that the opportunity cost of fighting is
lowered, thereby incentivizing participation [@Regan2005]. However,
this is predicated on the government lacking the capacity to fully
eradicate any rebel group. Where that capacity exists the state is
free to engage in reprisals.

It is at this point where the concept of executive constraints then
enters into the calculus. Leaders in nondemocratic states are more
likely to engage in violent repression as a response to internal
dissent [@Davenport2007a]. Executive constraints provide an
accountability mechanism deterring state violence by the leader
against their own citizens. Thus, where there are strong and effective
constraints on the executive, the opportunity cost for joining the
rebel side should be higher since the lack of indiscriminate violence
makes non-participation an attractive alternative. Rebel leaders will
then have to rely more heavily on economic selective benefits
[@Regan2005], creating a higher barrier for mobilization and
decreasing the risk of civil conflict onset. It should be noted
though, that where this relationship may not hold is in situations of
low intensity insurgencies involving predatory rebel groups solely
motivated by rent-seeking, or resource extraction, rather than
secession or government overthrow. The threshold for mobilization in
such scenarios is significantly lower than in full scale civil wars
since such rebel groups are not actively challenging the state for
control, meaning that rebel groups can initiate a conflict regardless
of the current level of state repression.

Taken together, none of the aforementioned mechanisms are mutually
exclusive, and it is beyond the scope of the present paper to properly
adjudicate between them; however, they do converge on a single
testable proposition. Namely that higher levels of executive
constraints should be associated with a lowered risk of civil conflict
onset. This proposition can be further extended: if state weakness is
a primary driver of civil conflict and the relative capacity of the
state is effected by the levels of executive constraints, as depicted
in the first theoretical pathway regarding executive looting, then
state capacity could have a mediating affect on the relationship
between executive constraints and civil conflict onset. Another
extension would be if executive constraints affect the mobilization
potential for rebel groups, then there could also be a confounding
effect from the relative threshold for mobilization. Finally, the
relationship could still be confounded by the overall opportunity for
rebellion --- the degree of executive constraints may not be
particularly relevant in relatively stable countries lacking the
conditions conducive to carrying out a rebellion. Regardless of
conditionalities, contrary results would indicate either that
executive constraints are indeterminate to the risk of civil conflict
or possibly that there is even a positive relationship to conflict
onset since constrained executives may be less able to effectively
deal with violent challenges due to risk aversion to violence and/or
from the slowdown of having to seek approval from a plurality of
stakeholders before committing to decisive action.

One final caveat, hitherto the relationship of executive constraints
to the notion of vertical accountability --- constraints on leaders by
free and open multiparty elections as well as an independent and
critical media --- has been left unspecified. This is not to diminish
the role of elections or the media; indeed, vertical accountability
can plausibly play a role in all of the aforementioned mechanisms. For
example, studies on state repression have emphasised the importance of
non-elites in being able to vote elected offices out of office as an
effective deterrent against negative sanctioning by leaders
[@BuenoDeMesquita2005; @Davenport2007]. But, elections occur only
intermittently and it is not an automatic given that voters will
punish an incumbent, especially when that repression is against an
oppressed minority community. Furthermore, while the media can expose
wrongdoing by government officials, legal punishment can only flow
through government bodies [@ODonnell1998]. Thus, executive constraints
are interesting in their own right and vertical accountability should
not confound any potential negative relationship to civil conflict
onset.

@@

I model civil conflict onset as a multilevel regression where onset is
a function of a set of predictors, including executive constraints and
an array of control variables, together with several varying
intercepts. The units of analysis are all independent country-years
from 1952 until 2013, as determined by data availability, with *I*
countries indexed by $i = 1 \dotsc I$ and *T* years indexed by $t = 1
\dotsc T$. The dependent variable, conflict onset, is a Bernoulli
distributed binary variable where 1 denotes the onset of at least one
new civil conflict in a given country-year. The chosen link function
transforming the linear regression output into the probability of
conflict onset, $\nu_{it}$, is the inverse logit such that $\nu_{it}
\in [0, 1]$. [^invlogit] The choice of a logit link function avoids
the truncation necessary for a linear probability model. Being a
sigmoid function it also ensures that the marginal contributions from
the predictors are reduced at the tails for higher and lower
probability values. That is to say the added value of a one unit
increase in a predictor is higher when the predicted probability is
already at 0.5 at the boundary between a positive and negative outcome
versus when it is at 0.95.

[^invlogit]: Technically, the probabilities in a logit function are on
    the open interval, $(0, 1)$, but due to overflow/underflow in
    floating point calculations done by computers the endpoints can be
    observed.

@@

where **X** is the predictors including the latent measure for
executive constraints, $\beta$ the vector of regression slopes, and
$\Zeta$ the *K* sets of varying intercepts indexed by $j_k$. All
predictors are lagged by one year to control for time endogeneity and
standardized by subtracting the mean and dividing by the standard
deviation to aid in computation as well as comparison of the posterior
$\beta$ estimates.

For the present analysis there are three sets of varying intercepts,
or random effects, included in the model. This includes two partially
pooled grouping factors, year and country, to account for
cross-sectional unmodeled heterogeneity between countries and over
time [@Shor2007]. Also included is the number of peace years since the
last civil conflict incidence per country to control for the lingering
endogenous effects of violence on the likelihood that additional
conflict breaks out.

@@
